---
title: ''
output: word_document
bibliography: references.bib
csl: global-change-biology.csl #(this will change depending on journal we choose. GCB would be a great choice)
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\raggedright

**Title:** 

**Authors (not yet complete, just alphabetical for now--final order TBD):** 

Cameron Dow^1^
Erika B. Gonzalez-Akre^1^,
Ryan Helcoski^1^,
Valentine Herrmann^1^,
Ian McGregor^1^,
Sean McMahon,
Albert Kim^1^,
Kristina J. Anderson-Teixeira^1,n^*



**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA

n. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama


*corresponding author: teixeirak@si.edu; +1 540 635 6546



```{r eval=FALSE, echo=FALSE}
## This is a chunk that won't be evaluated or displayed. I use these for notes.
## Target journal: ??? Global Change Biology????
## Instructions for authors: 
#[insert guidance on word count, n figures/ tables allowed, and then some rough planning of what goes in this manuscript]

```

\newpage

### Abstract



**Keywords**:  

\newpage

### Introduction

*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*

### Materials and Methods


### Results


### Discussion

### Acknowledgements
ForestGEO

### Authors' contributions

### References




