---
title: ''
output: word_document
bibliography: references.bib
csl: global-change-biology.csl #(this will change depending on journal we choose. GCB would be a great choice)
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\raggedright

**Title:** Effect of spring temperatures on tree growth phenology in a temperate deciduous forest

**Authors (not yet complete, just alphabetical for now--final order TBD):** 

Cameron Dow^1^
Erika B. Gonzalez-Akre^1^,
Ryan Helcoski^1^,
Valentine Herrmann^1^,
Ian McGregor^1^,
Sean McMahon,
Albert Kim^1^,
Kristina J. Anderson-Teixeira^1,n^*



**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA

n. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama


*corresponding author: teixeirak@si.edu; +1 540 635 6546



```{r eval=FALSE, echo=FALSE}
## This is a chunk that won't be evaluated or displayed. I use these for notes.
## Target journal: ??? Global Change Biology????
## Instructions for authors: 
#[insert guidance on word count, n figures/ tables allowed, and then some rough planning of what goes in this manuscript]

```

\newpage

### Abstract



**Keywords**:  

\newpage

### Introduction



**With climate warming, we can expect earlier springs in temperate forests.** 
These are likely to affect tree growth, woody productivity, and ecosystem C cycling.
However, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.

Here, we ...
- use 10 years of biweekly dendro band measurements for # species to characterize seasonal growth phenology 

- combine these records with 110 years of tree-ring data for 12 species to examine how spring temperatures impact annual tree growth

- examine the effects of an extremely early spring (2020) seasonal growth phenology and annual growth 

### Materials and Methods

#### Study site and data
SCBI ForestGEO plot

10 years biweekly dendrometer measurements

leaf phenology data?

- leaf phenology data from NEON and/or satellite-based (leaf phenology network: https://www.usanpn.org/news/spring) ?

- NDVI or PRI? https://onlinelibrary.wiley.com/doi/10.1111/gcb.15112 (from Ian)

- From Ian:  I noticed how the growth patterns observed by remote sensing pretty much mirror what Sean was finding with his dendro R package (btw is that functional on CRAN yet?) for both SERC and SCBI dendroband data.

perhaps bring in cores?

#### Analysis

*Here, I'll insert a reference to Sean's paper [@mcmahon_general_2015]. This is pulled from references.bib.*



### Results


### Discussion

*Content to incorporate:
N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact [@zohner_late-spring_2020]*

### Acknowledgements
ForestGEO

### Authors' contributions

### References




